---
title: "01 - Download"
author: "Darius Görgen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{01 - Download}
 %\VignetteEngine{knitr::rmarkdown}
 %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

## Preparation

The sen2tool package allows us to easily download and work with Sentinel-2 (S2) Level 1C datasets.
These datasets represent top of the atmosphere (TOA) reflectance values, thus they are not
suitable for every analysis. However, for most applications based on indices, they are able
to deliver robust results in a timely fashion. This is the main advantage of the present
package in comparison to other packages working with S2 data such as sen2r. sen2r works
on level 2B datasets which means that either you have to order the corresponding tiles
from the SentinelHub or process the level 1C data locally on your machine using sen2cor.
Both methods consume a lot of time and resources.
sen2tool downloads S2 data from a public Google Bucket which are available directly without
the need of a user account or waiting time for placing orders. To download from the bucket
one prerequisite is that the download tool gsutil is installed on your machine and available
on the command line. Instructions to install the tool on various operating systems (OS)
can be found [here](https://cloud.google.com/storage/docs/gsutil_install).
Once you installed the software open a command line of your choice and enter
the following line. You should get some version info on the gsutil tool otherwise
something went wrong during the installation. 

```{bash, eval = F}
gsutil version
```

Another option to use mapme.vegetation in a reproducible environment is to use the Docker engine
to create a containers based on our [Dockerfile](../../Dockerfile). Here, gsutil is automatically
installed when the image is created.

## Querying data

To query data, the first step is to load the necessary libraries.

```{r setup, eval = F}
library(sf)
library(stars)
library(mapme.vegetation)
```
For querying data there are two mandatory inputs. One is an area of interest represented
by a `sf` object the other is the time window for which we are querying the data. An optional
input is the maximum of cloud cover allowed for a scene to appear in our query. If it is not
specified, then all scenes matching the mandatory inputs are downloaded.

We will now load an example object which is found within the `sen2tool` package.

```{r read-in, eval = F}
aoi = st_read(system.file("extdata", "testregion.gpkg", package = "mapme.vegetation"))
```

There are two options you can decide how `mapme.vegetation` internally is querying matching 
scenes. The one uses less local storage (~1.6 GB) but is usually slower, while
the other occupies more local storage (~5 GB) but queries are faster retrieved.
When the package is freshly installed and the `downloadS2()` function is used for the
first time, you should determine your decision with the `use_db` option. This options
default is `TRUE` meaning that during the first execution the tile index file
is downloaded from Google Bucket and translated to a local SQLite data base.
Initially, this process takes some time, however as stated above future queries
are then processed more quickly. In case you opt for not using a local database 
(`use_db = FALSE`) the tile index is also downloaded and translated to a `.csv` file.
Queries are then applied to the `.csv` file which is why they are usually slower
compared to the database approach.
Note that either option you chose, the corresponding file is stored in the local
installation path of the `sen2tool` packages. This is why when the package is 
re-installed, the tile index needs to be downloaded and processed again. You
can find the local installation path of the package with `r find.package("sen2tool")`.

Now, let us send a first query.

```{r download, eval = F}
tiles_seasonal = downloadS2(aoi = aoi,
                            time_window = c("2016-06-01", "2019-09-30"),
                            time_period = "seasonal",
                            rootdir = ".",
                            extent_name = "test",
                            max_cloud = 75,
                            use_db = TRUE,
                            force_update = F,
                            query_only = T )
```

This query needs some explanation. First, we specified our area of interest (AOI)
which fixes the spatial extent for matching scenes. Second, we queried data
between the June, 1st 2016 and September, 30th 2019. However, with the option
`time_period = "seasonal"` we tell the function not to deliver every matching
scene between these two dates, but to deliver seasonal matches meaning that
every scene with a date between June, 1st and September 30th is returned for 
the years 2016 to 2019. The `rootdir` options specifies the directory to which
files are downloaded. They are then placed inside a child directory of `rootdir`
named according to the option `extent_name`. We also specified that the maximum cloud 
cover for matching scenes should not be higher than 75 %. With `use_db = TRUE`
we told the function to check for an SQLite database in the package directory and query
the data from there. We could have specified `force_update = TRUE` which would cause
the function to download the most recent tile index from the Google Bucket again and update the database.
Last, we told the function to not really download the data, but only apply the query.
This will return a data.frame with some information of the matching scenes.

Here we found 113 matching results for our query. Let's see how many scenes
will match our AOI when we change some of the parameters.

```{r download2, eval = F}
tiles_all = downloadS2(aoi = aoi,
                       time_window = c("2016-06-01", "2019-09-30"),
                       time_period = "full",
                       rootdir = ".",
                       extent_name = "test",
                       max_cloud = 20,
                       create_cloudmask = T,
                       use_db = TRUE,
                       force_update = F,
                       query_only = T)
```

We changed the maximum allowed cloud cover and changed the `time_period` option to "full".
In this case the query delivered 226 results.
When we leave out the `query_only` option or explicitly set it to `FALSE`, the matching data
will be downloaded into `<rootdir>/<extent_name>/L1C`. The function will print
some informative messages during the process, however, since the download is time consuming
this is not presented in this documentation. Note, that for this case where we actually
download the data we also enabled the option `create_cloudmask = TRUE`. This option
will translate the vector cloud mask which is delivered with the data to a GTiff
file within the `IMG_DATA` directory. This file is needed when calculating
vegetation indices to get cloud free images as explained in the next article on
[Calculation of Indices](E02-calcIndices.html).
